# Comparison of two PCA packages

-   Datasets: `decathlon2`

-   Algorithms:

    -   PCA

## prcomp vs princomp

<http://www.sthda.com/english/articles/31-principal-component-methods-in-r-practical-guide/118-principal-component-analysis-in-r-prcomp-vs-princomp/>

## General methods for principal component analysis

There are two general methods to perform PCA in R :

-   Spectral decomposition which examines the covariances / correlations between variables
-   Singular value decomposition which examines the covariances / correlations between individuals

The function princomp() uses the spectral decomposition approach. The functions prcomp() and PCA()[FactoMineR] use the singular value decomposition (SVD).

### prcomp() and princomp() functions

The simplified format of these 2 functions are :

    prcomp(x, scale = FALSE)
    princomp(x, cor = FALSE, scores = TRUE)

1.  Arguments for prcomp():\
    `x`: a numeric matrix or data frame\
    `scale`: a logical value indicating whether the variables should be scaled to have unit variance before the analysis takes place

2.  Arguments for princomp():\
    `x`: a numeric matrix or data frame `cor`: a logical value. If TRUE, the data will be centered and scaled before the analysis `scores`: a logical value. If TRUE, the coordinates on each principal component are calculated

### Loading factoextra

```{r eval=FALSE}
# install.packages("factoextra")
```

```{r}
library(factoextra)
```

### Loading the decathlon dataset

We'll use the data sets `decathlon2` [in factoextra], which has been already described at: PCA - Data format.

Briefly, it contains:

-   Active individuals (rows 1 to 23) and active variables (columns 1 to 10), which are used to perform the principal component analysis
-   Supplementary individuals (rows 24 to 27) and supplementary variables (columns 11 to 13), which coordinates will be predicted using the PCA information and parameters obtained with active individuals/variables.

```{r}
library("factoextra")
data(decathlon2)
decathlon2.active <- decathlon2[1:23, 1:10]
head(decathlon2.active[, 1:6])
```

```{r}
decathlon2.supplementary <- decathlon2[24:27, 1:10]
head(decathlon2.supplementary[, 1:6])
```

## Compute PCA in R using prcomp()

In this section we'll provide an easy-to-use R code to compute and visualize PCA in R using the prcomp() function and the factoextra package.

\`. Load factoextra for visualization

```{r}
library(factoextra)
```

2.  compute PCA

```{r}
# compute PCA
res.pca <- prcomp(decathlon2.active, scale = TRUE)
```

3.  Visualize eigenvalues (scree plot). Show the percentage of variances explained by each principal component.

```{r}
# Visualize eigenvalues (scree plot).
fviz_eig(res.pca, addlabels = TRUE, ylim = c(0, 50))
```

> From the plot above, we might want to stop at the fifth principal component. 87% of the information (variances) contained in the data are retained by the first five principal components.

## Plots: quality and contribution

4.  Graph of individuals. Individuals with a similar profile are grouped together.

```{r}
# Graph of individuals.
fviz_pca_ind(res.pca,
             col.ind = "cos2", # Color by the quality of representation
             gradient.cols = c("#00AFBB", "#E7B800", "#FC4E07"),
             repel = TRUE     # Avoid text overlapping
             )
```

5.  Graph of variables. Positive correlated variables point to the same side of the plot. Negative correlated variables point to opposite sides of the graph.

```{r}
# Graph of variables.
fviz_pca_var(res.pca,
             col.var = "contrib", # Color by contributions to the PC
             gradient.cols = c("#00AFBB", "#E7B800", "#FC4E07"),
             repel = TRUE     # Avoid text overlapping
             )
```

6.  Biplot of individuals and variables

```{r}
# Biplot of individuals and variables
fviz_pca_biplot(res.pca, repel = TRUE,
                col.var = "#2E9FDF", # Variables color
                col.ind = "#696969"  # Individuals color
                )
```

## Access to the PCA results

```{r}
library(factoextra)
# Eigenvalues
eig.val <- get_eigenvalue(res.pca)
eig.val
  
# Results for Variables
res.var <- get_pca_var(res.pca)
res.var$coord          # Coordinates
res.var$contrib        # Contributions to the PCs
res.var$cos2           # Quality of representation 
# Results for individuals
res.ind <- get_pca_ind(res.pca)
res.ind$coord          # Coordinates
res.ind$contrib        # Contributions to the PCs
res.ind$cos2           # Quality of representation 
```

## Predict using PCA

In this section, we'll show how to predict the coordinates of supplementary individuals and variables using only the information provided by the previously performed PCA.

1.  Data: rows 24 to 27 and columns 1 to to 10 [in decathlon2 data sets]. The new data must contain columns (variables) with the same names and in the same order as the active data used to compute PCA.

```{r}
# Data for the supplementary individuals
ind.sup <- decathlon2[24:27, 1:10]
ind.sup[, 1:6]
```

2.  Predict the coordinates of new individuals data. Use the R base function predict():

```{r}
ind.sup.coord <- predict(res.pca, newdata = ind.sup)
ind.sup.coord[, 1:4]
```

3.  Graph of individuals including the supplementary individuals:

```{r}
# Plot of active individuals
p <- fviz_pca_ind(res.pca, repel = TRUE)
# Add supplementary individuals
fviz_add(p, ind.sup.coord, color ="blue")
```

The predicted coordinates of individuals can be manually calculated as follow:

1.  Center and scale the new individuals data using the center and the scale of the PCA
2.  Calculate the predicted coordinates by multiplying the scaled values with the eigenvectors (loadings) of the principal components. The R code below can be used :

```{r}
# Centering and scaling the supplementary individuals
ind.scaled <- scale(ind.sup, 
                    center = res.pca$center,
                    scale = res.pca$scale)
# Coordinates of the individividuals
coord_func <- function(ind, loadings){
  r <- loadings*ind
  apply(r, 2, sum)
}
pca.loadings <- res.pca$rotation
ind.sup.coord <- t(apply(ind.scaled, 1, coord_func, pca.loadings ))
ind.sup.coord[, 1:4]
```

## Supplementary variables

### Qualitative / categorical variables

The data sets decathlon2 contain a supplementary qualitative variable at columns 13 corresponding to the type of competitions.

Qualitative / categorical variables can be used to color individuals by groups. The grouping variable should be of same length as the number of active individuals (here 23).

```{r}
groups <- as.factor(decathlon2$Competition[1:23])
fviz_pca_ind(res.pca,
             col.ind = groups, # color by groups
             palette = c("#00AFBB",  "#FC4E07"),
             addEllipses = TRUE, # Concentration ellipses
             ellipse.type = "confidence",
             legend.title = "Groups",
             repel = TRUE
             )
```

Calculate the coordinates for the levels of grouping variables. The coordinates for a given group is calculated as the mean coordinates of the individuals in the group.

```{r}
library(magrittr) # for pipe %>%
library(dplyr)   # everything else

# 1. Individual coordinates
res.ind <- get_pca_ind(res.pca)
# 2. Coordinate of groups
coord.groups <- res.ind$coord %>%
  as_data_frame() %>%
  select(Dim.1, Dim.2) %>%
  mutate(competition = groups) %>%
  group_by(competition) %>%
  summarise(
    Dim.1 = mean(Dim.1),
    Dim.2 = mean(Dim.2)
    )
coord.groups
```

### Quantitative variables

Data: columns 11:12. Should be of same length as the number of active individuals (here 23)

```{r}
quanti.sup <- decathlon2[1:23, 11:12, drop = FALSE]
head(quanti.sup)
```

The coordinates of a given quantitative variable are calculated as the correlation between the quantitative variables and the principal components.

```{r}
# Predict coordinates and compute cos2
quanti.coord <- cor(quanti.sup, res.pca$x)
quanti.cos2 <- quanti.coord^2
# Graph of variables including supplementary variables
p <- fviz_pca_var(res.pca)
fviz_add(p, quanti.coord, color ="blue", geom="arrow")
```

## Theory behind PCA results

### PCA results for variables

Here we'll show how to calculate the PCA results for variables: coordinates, cos2 and contributions:

`var.coord` = loadings \* the component standard deviations `var.cos2` = var.coord\^2 `var.contrib`. The contribution of a variable to a given principal component is (in percentage) : (var.cos2 \* 100) / (total cos2 of the component)

```{r}
# Helper function 
#::::::::::::::::::::::::::::::::::::::::
var_coord_func <- function(loadings, comp.sdev){
  loadings*comp.sdev
}
```

```{r}
# Compute Coordinates
#::::::::::::::::::::::::::::::::::::::::
loadings <- res.pca$rotation
sdev <- res.pca$sdev
var.coord <- t(apply(loadings, 1, var_coord_func, sdev)) 
head(var.coord[, 1:4])
```

```{r}
# Compute Cos2
#::::::::::::::::::::::::::::::::::::::::
var.cos2 <- var.coord^2
head(var.cos2[, 1:4])
```

```{r}
# Compute contributions
#::::::::::::::::::::::::::::::::::::::::
comp.cos2 <- apply(var.cos2, 2, sum)
contrib <- function(var.cos2, comp.cos2){var.cos2*100/comp.cos2}
var.contrib <- t(apply(var.cos2,1, contrib, comp.cos2))
head(var.contrib[, 1:4])
```

### PCA results for individuals

-   `ind.coord` = res.pca\$x

-   Cos2 of individuals. Two steps:

    -   Calculate the square distance between each individual and the PCA center of gravity: d2 = [(var1_ind_i - mean_var1)/sd_var1]\^2 + ...+ [(var10_ind_i - mean_var10)/sd_var10]\^2 + ...+..
    -   Calculate the cos2 as ind.coord\^2/d2

-   Contributions of individuals to the principal components: 100 \* (1 / number_of_individuals)\*(ind.coord\^2 / comp_sdev\^2). Note that the sum of all the contributions per column is 100

```{r}
# Coordinates of individuals
#::::::::::::::::::::::::::::::::::
ind.coord <- res.pca$x
head(ind.coord[, 1:4])
```

```{r}
# Cos2 of individuals
#:::::::::::::::::::::::::::::::::
# 1. square of the distance between an individual and the
# PCA center of gravity
center <- res.pca$center
scale<- res.pca$scale

getdistance <- function(ind_row, center, scale){
  return(sum(((ind_row-center)/scale)^2))
}

d2 <- apply(decathlon2.active,1, getdistance, center, scale)
# 2. Compute the cos2. The sum of each row is 1
cos2 <- function(ind.coord, d2){return(ind.coord^2/d2)}
ind.cos2 <- apply(ind.coord, 2, cos2, d2)
head(ind.cos2[, 1:4])
```

```{r}
# Contributions of individuals
#:::::::::::::::::::::::::::::::
contrib <- function(ind.coord, comp.sdev, n.ind){
  100*(1/n.ind)*ind.coord^2/comp.sdev^2
}
ind.contrib <- t(apply(ind.coord, 1, contrib, 
                       res.pca$sdev, nrow(ind.coord)))
head(ind.contrib[, 1:4])
```
