---
editor_options: 
  markdown: 
    wrap: 72
---

# (PART) Feature Engineering {.unnumbered}

# Ten methods to assess Variable Importance

-   Datasets: `GlaucomaM`

-   Algorithms:

    -   Partition Trees
    -   Regularized Random Forest (RRF)
    -   Lasso Regression
    -   Linear Regression
    -   Recursive Feature Elimination (RFE)
    -   Genetic Algorithm
    -   Simulated Annealing

Source:
<https://www.machinelearningplus.com/machine-learning/feature-selection/>

In real-world datasets, it is fairly common to have columns that are
nothing but noise.

You are better off getting rid of such variables because of the memory
space they occupy, the time and the computational esources it is going
to cost, especially in large datasets.

Sometimes, you have a variable that makes business sense, but you are
not sure if it actually helps in predicting the Y. You also need to
consider the fact that, a feature that could be useful in one ML
algorithm (say a decision tree) may go underrepresented or unused by
another (like a regression model).

Having said that, it is still possible that a variable that shows poor
signs of helping to explain the response variable (Y), can turn out to
be significantly useful in the presence of (or combination with) other
predictors. What I mean by that is, a variable might have a low
correlation value of (\~0.2) with Y. But in the presence of other
variables, it can help to explain certain patterns/phenomenon that other
variables can't explain.

In such cases, it can be hard to make a call whether to include or
exclude such variables.

The strategies we are about to discuss can help fix such problems. Not
only that, it will also help understand if a particular variable is
important or not and how much it is contributing to the model

An important caveat. It is always best to have variables that have sound
business logic backing the inclusion of a variable and rely solely on
variable importance metrics.

Alright. Let's load up the 'Glaucoma' dataset where the goal is to
predict if a patient has Glaucoma or not based on 63 different
physiological measurements. You can directly run the codes or download
the dataset here.

A lot of interesting examples ahead. Let's get started.

```{r}
# Load Packages and prepare dataset
library(DALEX)
library(TH.data)
library(caret)
library(tictoc)

data("GlaucomaM", package = "TH.data")
trainData <- GlaucomaM
head(trainData)
```

## 1. Boruta

Boruta is a feature ranking and selection algorithm based on random
forests algorithm. The advantage with Boruta is that it clearly decides
if a variable is important or not and helps to select variables that are
statistically significant. Besides, you can adjust the strictness of the
algorithm by adjusting the $p$ values that defaults to 0.01 and the
`maxRuns`.

`maxRuns` is the number of times the algorithm is run. The higher the
`maxRuns` the more selective you get in picking the variables. The
default value is 100.

In the process of deciding if a feature is important or not, some
features may be marked by Boruta as 'Tentative'. Sometimes increasing
the maxRuns can help resolve the 'Tentativeness' of the feature.

Lets see an example based on the Glaucoma dataset from `TH.data` package
that I created earlier.

```{r}
# install.packages('Boruta')
library(Boruta)
```

The `boruta` function uses a formula interface just like most predictive
modeling functions. So the first argument to `boruta()` is the formula
with the response variable on the left and all the predictors on the
right. By placing a dot, all the variables in `trainData` other than
Class will be included in the model.

The `doTrace` argument controls the amount of output printed to the
console. Higher the value, more the log details you get. So save space I
have set it to 0, but try setting it to 1 and 2 if you are running the
code.

Finally the output is stored in `boruta_output`.

```{r train_boruta}
# Perform Boruta search
boruta_output <- Boruta(Class ~ ., data=na.omit(trainData), doTrace=0)  
```

Let's see what the boruta_output contains.

```{r}
names(boruta_output)
```

```{r}
# Get significant variables including tentatives
boruta_signif <- getSelectedAttributes(boruta_output, withTentative = TRUE)
print(boruta_signif)  
```

If you are not sure about the tentative variables being selected for
granted, you can choose a `TentativeRoughFix` on `boruta_output`.

```{r}
# Do a tentative rough fix
roughFixMod <- TentativeRoughFix(boruta_output)
boruta_signif <- getSelectedAttributes(roughFixMod)
print(boruta_signif)
```

There you go. Boruta has decided on the 'Tentative' variables on our
behalf. Let's find out the importance scores of these variables.

```{r}
# Variable Importance Scores
imps <- attStats(roughFixMod)
imps2 = imps[imps$decision != 'Rejected', c('meanImp', 'decision')]
head(imps2[order(-imps2$meanImp), ])  # descending sort
```

Let's plot it to see the importances of these variables.

```{r boruta-plot}
# Plot variable importance
plot(boruta_output, cex.axis=.7, las=2, xlab="", main="Variable Importance")  
```

This plot reveals the importance of each of the features.

The columns in green are 'confirmed' and the ones in red are not. There
are couple of blue bars representing `ShadowMax` and `ShadowMin.` They
are not actual features, but are used by the `boruta` algorithm to
decide if a variable is important or not.

## 2. Variable Importance

Another way to look at feature selection is to consider variables most
used by various ML algorithms the most to be important.

Depending on how the machine learning algorithm learns the relationship
between $X$'s and \$Y\$, different machine learning algorithms may
possibly end up using different variables (but mostly common vars) to
various degrees.

What I mean by that is, the variables that proved useful in a tree-based
algorithm like `rpart`, can turn out to be less useful in a
regression-based model. **So all variables need not be equally useful to
all algorithms.**

So how do we find the variable importance for a given ML algo?

`train()` the desired model using the caret package. Then, use
`varImp()` to determine the feature importance.

You may want to try out multiple algorithms, to get a feel of the
usefulness of the features across algos.

## 3. rpart

```{r rpart-setup}
# Train an rpart model and compute variable importance.
library(caret)
set.seed(100)
rPartMod <- train(Class ~ ., 
                  data=trainData, 
                  method="rpart")

rpartImp <- varImp(rPartMod)
print(rpartImp)
```

Only 5 of the 63 features was used by rpart and if you look closely, the
5 variables used here are in the top 6 that boruta selected.

Let's do one more: the variable importances from Regularized Random
Forest (RRF) algorithm.

## 4. Regularized Random Forest (RRF)

```{r rrf-train}
tic()
# Train an RRF model and compute variable importance.
set.seed(100)
rrfMod <- train(Class ~ ., 
                data = trainData, 
                method = "RRF")

rrfImp <- varImp(rrfMod, scale=F)
toc()
rrfImp
```

```{r}
plot(rrfImp, top = 20, main='Variable Importance')
```

The topmost important variables are pretty much from the top tier of
Boruta's selections.

Some of the other algorithms available in `train()` that you can use to
compute varImp are the following:

    ada, AdaBag, AdaBoost.M1, adaboost, bagEarth, bagEarthGCV, bagFDA, bagFDAGCV, bartMachine, blasso, BstLm, bstSm, C5.0, C5.0Cost, C5.0Rules, C5.0Tree, cforest, chaid, ctree, ctree2, cubist, deepboost, earth, enet, evtree, extraTrees, fda, gamboost, gbm_h2o, gbm, gcvEarth, glmnet_h2o, glmnet, glmStepAIC, J48, JRip, lars, lars2, lasso, LMT, LogitBoost, M5, M5Rules, msaenet, nodeHarvest, OneR, ordinalNet, ORFlog, ORFpls, ORFridge, ORFsvm, pam, parRF, PART, penalized, PenalizedLDA, qrf, ranger, Rborist, relaxo, rf, rFerns, rfRules, rotationForest, rotationForestCp, rpart, rpart1SE, rpart2, rpartCost, rpartScore, rqlasso, rqnc, RRF, RRFglobal, sdwd, smda, sparseLDA, spikeslab, wsrf, xgbLinear, xgbTree.

## 5. Lasso Regression

Least Absolute Shrinkage and Selection Operator (LASSO) regression is a
type of regularization method that penalizes with L1-norm.

It basically imposes a cost to having large weights (value of
coefficients). And its called L1 regularization, because the cost added,
is proportional to the absolute value of weight coefficients.

As a result, in the process of shrinking the coefficients, it eventually
reduces the coefficients of certain unwanted features all the to zero.
That is, it removes the unneeded variables altogether.

So effectively, LASSO regression can be considered as a variable
selection technique as well.

```{r lasso-train}
library(glmnet)

# online data
# trainData <- read.csv('https://raw.githubusercontent.com/selva86/datasets/master/GlaucomaM.csv')

trainData <- read.csv(file.path(data_raw_dir, "glaucoma.csv"))

x <- as.matrix(trainData[,-63]) # all X vars
y <- as.double(as.matrix(ifelse(trainData[, 63]=='normal', 0, 1))) # Only Class

# Fit the LASSO model (Lasso: Alpha = 1)
set.seed(100)
cv.lasso <- cv.glmnet(x, y, family='binomial', alpha=1, parallel=TRUE, standardize=TRUE, type.measure='auc')

# Results
plot(cv.lasso)
```

Let's see how to interpret this plot.

The X axis of the plot is the log of `lambda`. That means when it is 2
here, the lambda value is actually 100.

The numbers at the top of the plot show how many predictors were
included in the model. The position of red dots along the Y-axis tells
what `AUC` we got when you include as many variables shown on the top
x-axis.

You can also see two dashed vertical lines.

The first one on the left points to the lambda with the lowest mean
squared error. The one on the right point to the number of variables
with the highest deviance within 1 standard deviation.

The best lambda value is stored inside 'cv.lasso\$lambda.min'.

```{r}
# plot(cv.lasso$glmnet.fit, xvar="lambda", label=TRUE)
cat('Min Lambda: ', cv.lasso$lambda.min, '\n 1Sd Lambda: ', cv.lasso$lambda.1se)
df_coef <- round(as.matrix(coef(cv.lasso, s=cv.lasso$lambda.min)), 2)

# See all contributing variables
df_coef[df_coef[, 1] != 0, ]
```

The above output shows what variables LASSO considered important. A high
positive or low negative implies more important is that variable.

## 6. Step wise Forward and Backward Selection

Stepwise regression can be used to select features if the Y variable is
a numeric variable. It is particularly used in selecting best linear
regression models.

It searches for the best possible regression model by iteratively
selecting and dropping variables to arrive at a model with the lowest
possible AIC.

It can be implemented using the `step()` function and you need to
provide it with a lower model, which is the base model from which it
won't remove any features and an upper model, which is a full model that
has all possible features you want to have.

Our case is not so complicated (\< 20 vars), so lets just do a simple
stepwise in 'both' directions.

I will use the `ozone` dataset for this where the objective is to
predict the `ozone_reading` based on other weather related observations.

```{r data-ozone1}
# Load data
# online
# trainData <- read.csv("http://rstatistics.net/wp-content/uploads/2015/09/ozone1.csv",
#                      stringsAsFactors=F)
trainData <- read.csv(file.path(data_raw_dir, "ozone1.csv"))
print(head(trainData))
```

The data is ready. Let's perform the stepwise.

```{r step-wise}
# Step 1: Define base intercept only model
base.mod <- lm(ozone_reading ~ 1 , data=trainData)  

# Step 2: Full model with all predictors
all.mod <- lm(ozone_reading ~ . , data= trainData) 

# Step 3: Perform step-wise algorithm. direction='both' implies both forward and backward stepwise
stepMod <- step(base.mod, scope = list(lower = base.mod, upper = all.mod), direction = "both", trace = 0, steps = 1000)  

# Step 4: Get the shortlisted variable.
shortlistedVars <- names(unlist(stepMod[[1]])) 
shortlistedVars <- shortlistedVars[!shortlistedVars %in% "(Intercept)"] # remove intercept

# Show
print(shortlistedVars)
```

The selected model has the above 6 features in it.

But if you have too many features (\> 100) in training data, then it
might be a good idea to split the dataset into chunks of 10 variables
each with Y as mandatory in each dataset. Loop through all the chunks
and collect the best features.

We are doing it this way because some variables that came as important
in a training data with fewer features may not show up in a linear reg
model built on lots of features.

Finally, from a pool of shortlisted features (from small chunk models),
run a full stepwise model to get the final set of selected features.

You can take this as a learning assignment to be solved within 20
minutes.

## 7. Relative Importance from Linear Regression

This technique is specific to linear regression models.

Relative importance can be used to assess which variables contributed
how much in explaining the linear model's R-squared value. So, if you
sum up the produced importances, it will add up to the model's R-sq
value.

In essence, it is not directly a feature selection method, because you
have already provided the features that go in the model. But after
building the model, the `relaimpo` can provide a sense of how important
each feature is in contributing to the R-sq, or in other words, in
'explaining the Y variable'.

So, how to calculate relative importance?

It is implemented in the `relaimpo` package. Basically, you build a
linear regression model and pass that as the main argument to
`calc.relimp()`. relaimpo has multiple options to compute the relative
importance, but the recommended method is to use `type='lmg'`, as I have
done below.

```{r relaimpo-ozone}
# install.packages('relaimpo')
library(relaimpo)

# Build linear regression model
model_formula = ozone_reading ~ Temperature_Sandburg + Humidity + Temperature_ElMonte + Month + pressure_height + Inversion_base_height
lmMod <- lm(model_formula, data=trainData)

# calculate relative importance
relImportance <- calc.relimp(lmMod, type = "lmg", rela = F)  

# Sort
cat('Relative Importances: \n')
sort(round(relImportance$lmg, 3), decreasing=TRUE)
```

Additionally, you can use bootstrapping (using `boot.relimp`) to compute
the confidence intervals of the produced relative importances.

```{r}
bootsub <- boot.relimp(ozone_reading ~ Temperature_Sandburg + Humidity + Temperature_ElMonte + Month + pressure_height + Inversion_base_height, data=trainData,
                       b = 1000, type = 'lmg', rank = TRUE, diff = TRUE)

plot(booteval.relimp(bootsub, level=.95))
```

## 8. Recursive Feature Elimination (RFE)

Recursive feature elimnation (rfe) offers a rigorous way to determine
the important variables before you even feed them into a ML algo.

It can be implemented using the `rfe()` from caret package.

The rfe() also takes two important parameters.

-   `sizes`
-   `rfeControl`

So, what does `sizes` and `rfeControl` represent?

The sizes determines the number of most important features the rfe
should iterate. Below, I have set the size as 1 to 5, 10, 15 and 18.

Secondly, the `rfeControl` parameter receives the output of the
`rfeControl()`. You can set what type of variable evaluation algorithm
must be used. Here, I have used random forests based rfFuncs. The
`method='repeatedCV'` means it will do a repeated k-Fold cross
validation with `repeats=5`.

Once complete, you get the accuracy and kappa for each model size you
provided. The final selected model subset size is marked with a \* in
the rightmost selected column.

```{r}
str(trainData)
```

```{r rfe}
tic()
set.seed(100)
options(warn=-1)

subsets <- c(1:5, 10, 15, 18)

ctrl <- rfeControl(functions = rfFuncs,
                   method = "repeatedcv",
                   repeats = 5,
                   verbose = FALSE)

lmProfile <- rfe(x=trainData[, c(1:3, 5:13)], y=trainData$ozone_reading,
                 sizes = subsets,
                 rfeControl = ctrl)
toc()
lmProfile
```

So, it says, Temperature_ElMonte, Pressure_gradient,
Temperature_Sandburg, Inversion_temperature, Humidity are the top 5
variables in that order.

And the best model size out of the provided models sizes (in subsets) is
10.

You can see all of the top 10 variables from '`lmProfile$optVariables`'
that was created using `rfe` function above.

## 9. Genetic Algorithm

You can perform a supervised feature selection with genetic algorithms
using the `gafs()`. This is **quite resource expensive** so consider
that before choosing the number of iterations (iters) and the number of
repeats in `gafsControl()`.

```{r Genetic_Algorithm, warning=FALSE, message=FALSE}
tic()
# Define control function
ga_ctrl <- gafsControl(functions = rfGA,  # another option is `caretGA`.
                        method = "cv",
                        repeats = 3)

# Genetic Algorithm feature selection
set.seed(100)
ga_obj <- gafs(x=trainData[, c(1:3, 5:13)], 
               y=trainData[, 4], 
               iters = 3,   # normally much higher (100+)
               gafsControl = ga_ctrl)
toc()
ga_obj
```

```{r}
# Optimal variables
ga_obj$optVariables
```

'Month' 'Day_of_month' 'Wind_speed' 'Temperature_ElMonte'
'Pressure_gradient' 'Visibility'

So the optimal variables according to the genetic algorithms are listed
above. But, I wouldn't use it just yet because, the above variant was
tuned for only 3 iterations, which is quite low. I had to set it so low
to save computing time.

## 10. Simulated Annealing

Simulated annealing is a global search algorithm that allows a
suboptimal solution to be accepted in hope that a better solution will
show up eventually.

It works by making small random changes to an initial solution and sees
if the performance improved. The change is accepted if it improves, else
it can still be accepted if the difference of performances meet an
acceptance criteria.

In caret it has been implemented in the `safs()` which accepts a control
parameter that can be set using the `safsControl()` function.

`safsControl` is similar to other control functions in caret (like you
saw in rfe and ga), and additionally it accepts an improve parameter
which is the number of iterations it should wait without improvement
until the values are reset to previous iteration.

```{r Simulated_Annealing}
tic()
# Define control function
sa_ctrl <- safsControl(functions = rfSA,
                        method = "repeatedcv",
                        repeats = 3,
                        improve = 5) # n iterations without improvement before a reset

# Genetic Algorithm feature selection
set.seed(100)
sa_obj <- safs(x=trainData[, c(1:3, 5:13)], 
               y=trainData[, 4],
               safsControl = sa_ctrl)
toc()
sa_obj
```

```{r}
# Optimal variables
print(sa_obj$optVariables)
```

## Information Value and Weights of Evidence

The Information Value can be used to judge how important a given
categorical variable is in explaining the binary Y variable. It goes
well with logistic regression and other classification models that can
model binary variables.

Let's try to find out how important the categorical variables are in
predicting if an individual will earn \> 50k from the `adult.csv`
dataset. Just run the code below to import the dataset.

```{r infovalue}
library(InformationValue)

# online data
# inputData <- read.csv("http://rstatistics.net/wp-content/uploads/2015/09/adult.csv")

inputData <- read.csv(file.path(data_raw_dir, "adult.csv"))
print(head(inputData))
```

```{r get_information_value}
# Choose Categorical Variables to compute Info Value.
cat_vars <- c ("WORKCLASS", "EDUCATION", "MARITALSTATUS", "OCCUPATION", "RELATIONSHIP", "RACE", "SEX", "NATIVECOUNTRY")  # get all categorical variables

factor_vars <- cat_vars


# Init Output
df_iv <- data.frame(VARS=cat_vars, IV=numeric(length(cat_vars)), STRENGTH=character(length(cat_vars)), stringsAsFactors = F)  # init output dataframe

# Get Information Value for each variable
for (factor_var in factor_vars){
  df_iv[df_iv$VARS == factor_var, "IV"] <- InformationValue::IV(X=inputData[, factor_var], Y=inputData$ABOVE50K)
  df_iv[df_iv$VARS == factor_var, "STRENGTH"] <- attr(InformationValue::IV(X=inputData[, factor_var], Y=inputData$ABOVE50K), "howgood")
}

# Sort
df_iv <- df_iv[order(-df_iv$IV), ]

df_iv
```

Here is what the quantum of Information Value means:

Less than 0.02, then the predictor is not useful for modeling
(separating the Goods from the Bads)

0.02 to 0.1, then the predictor has only a weak relationship. 0.1 to
0.3, then the predictor has a medium strength relationship. 0.3 or
higher, then the predictor has a strong relationship. That was about IV.
Then what is Weight of Evidence?

Weights of evidence can be useful to find out how important a given
categorical variable is in explaining the 'events' (called 'Goods' in
below table.)

The 'Information Value' of the categorical variable can then be derived
from the respective WOE values.

IV=(perc good of all goods−perc bad of all bads) \*WOE

The 'WOETable' below given the computation in more detail.

```{r WOETable}
WOETable(X=inputData[, 'WORKCLASS'], Y=inputData$ABOVE50K)
```

The total IV of a variable is the sum of IV's of its categories.

## DALEX Package

The `DALEX` is a powerful package that explains various things about the
variables used in an ML model.

For example, using the `variable_dropout()` function you can find out
how important a variable is based on a dropout loss, that is how much
loss is incurred by removing a variable from the model.

Apart from this, it also has the `single_variable()` function that gives
you an idea of how the model's output will change by changing the values
of one of the X's in the model.

It also has the `single_prediction()` that can decompose a single model
prediction so as to understand which variable caused what effect in
predicting the value of Y.

```{r dalex-adult}
library(randomForest)
library(DALEX)

# Load data
# inputData <- read.csv("http://rstatistics.net/wp-content/uploads/2015/09/adult.csv")

inputData <- read.csv(file.path(data_raw_dir, "adult.csv"))

# Train random forest model
rf_mod <- randomForest(factor(ABOVE50K) ~ ., data = inputData, ntree=100)
rf_mod

# Variable importance with DALEX
explained_rf <- explain(rf_mod, data=inputData, y=inputData$ABOVE50K)

# Get the variable importances
# varimps = variable_dropout(explained_rf, type='raw')
varimps = variable_importance(explained_rf, type='raw')
print(varimps)
```

```{r}
plot(varimps)
```

## Conclusion

Hope you find these methods useful. As it turns out different methods
showed different variables as important, or at least the degree of
importance changed. This need not be a conflict, because each method
gives a different perspective of how the variable can be useful
depending on how the algorithms learn `Y ~ x`. So its cool.

If you find any code breaks or bugs, report the issue here or just write
it below.
