# Introduction to algorithms for Classification

* Datasets: `PimaIndiansDiabetes`

## Comparison of *CART, LDA, SVM, KNN, RF*


## Introduction
We compare the following classification algorithms:

* CART
* LDA
* SVM
* KNN
* RF

## Workflow

1. Load dataset
2. Create the train dataset
3. Train the models
4. Collect resamples
5. Plot comparison
6. Summarize p-values


```{r warning=FALSE, message=FALSE}
# load packages
library(mlbench)
library(caret)
# load the dataset
data(PimaIndiansDiabetes)
```

```{r}
dplyr::glimpse(PimaIndiansDiabetes)
```

```{r}
tibble::as_tibble(PimaIndiansDiabetes)
```

## Train the models using cross-validation

```{r set-train-control}
# prepare training scheme
trainControl <- trainControl(method = "repeatedcv", 
                             number=10, 
                             repeats=3)
```

```{r run-cart}
# CART
set.seed(7)
fit.cart <- train(diabetes~., data=PimaIndiansDiabetes, 
                  method = "rpart", trControl=trainControl)
```

```{r run-lda}
# LDA: Linear Discriminant Analysis
set.seed(7)
fit.lda <- train(diabetes~., data=PimaIndiansDiabetes, 
                 method="lda", trControl=trainControl)
```

```{r run-svm}
# SVM
set.seed(7)
fit.svm <- train(diabetes~., data=PimaIndiansDiabetes, 
                 method="svmRadial", trControl=trainControl)
```

```{r run-knn}
# KNN
set.seed(7)
fit.knn <- train(diabetes~., data=PimaIndiansDiabetes, 
                 method="knn", trControl=trainControl)
```

```{r run-random_forest-takes_a_while}
# Random Forest
set.seed(7)
fit.rf <- train(diabetes~., data=PimaIndiansDiabetes, 
                method="rf", trControl=trainControl)
```


```{r resample_results}
# collect resamples
results <- resamples(list(CART=fit.cart, 
                          LDA=fit.lda, 
                          SVM=fit.svm, 
                          KNN=fit.knn, 
                          RF=fit.rf))
```

## Compare models

```{r compare}
# summarize differences between models
summary(results)
```

## Plot comparison

```{r}
# box and whisker plots to compare models
scales <- list(x=list(relation="free"), y=list(relation="free"))
bwplot(results, scales=scales)
```

```{r density_plot, fig.asp=1}
# density plots of accuracy
scales <- list(x=list(relation="free"), y=list(relation="free"))
densityplot(results, scales=scales, pch = "|")
```

```{r dot_plot, fig.asp=1}
# dot plots of accuracy
scales <- list(x=list(relation="free"), y=list(relation="free"))
dotplot(results, scales=scales)
```


```{r parallel_plot, fig.asp=1}
# parallel plots to compare models
parallelplot(results)
```


```{r pairwise_plot, fig.asp=1}
# pairwise scatter plots of predictions to compare models
splom(results)
```

```{r xy_plot, fig.asp=1}
# xyplot plots to compare models
xyplot(results, models=c("LDA", "SVM"))
```

```{r differences_for_p_values}
# difference in model predictions
diffs <- diff(results)
# summarize p-values for pairwise comparisons
summary(diffs)
```


