# Comparing Classification algorithms 

* Datasets: `iris`
* Algorithms: *LDA, CART, KNN, SVM, RF*



## Introduction
These are the algorithms used:

1. LDA
2. CART
3. KNN
4. SVM
5. RF

```{r}
# load the caret package
library(caret)
# attach the iris dataset to the environment
data(iris)
# rename the dataset
dataset <- iris
```

## Workflow

1. Load dataset 
2. Create train and test datasets, 80/20
3. Inspect dataset
4. Visualize features
5. Set the train control to
  * 10 cross-validations
  * Metric: accuracy
6. Train the models 
7. Compare accuracy of models
8. Visual comparison
9. Make predictions on `validation` set


We will split the loaded dataset into two, $80\%$ of which we will use to train our models and $20\%$ that we will hold back as a validation dataset.

```{r}
# create a list of 80% of the rows in the original dataset we can use for training
validationIndex <- createDataPartition(dataset$Species, p=0.80, list=FALSE)
# select 20% of the data for validation
validation <- dataset[-validationIndex,]

# use the remaining 80% of data to training and testing the models
dataset <- dataset[validationIndex,]

# dimensions of dataset
dim(dataset)
```


```{r}
# list types for each attribute
sapply(dataset, class)
```

## Peek at the dataset

```{r}
# take a peek at the first 5 rows of the data
head(dataset)
```


```{r}
library(dplyr)

glimpse(dataset)
```

```{r}
library(skimr)
print(skim(dataset))
```

## Levels of the class

```{r}
# list the levels for the class
levels(dataset$Species)
```

## class distribution

```{r}
# summarize the class distribution
percentage <- prop.table(table(dataset$Species)) * 100
cbind(freq=table(dataset$Species), percentage=percentage)
```

## Visualize the dataset

```{r}
# split input and output
x <- dataset[,1:4]
y <- dataset[,5]
```


```{r}
# boxplot for each attribute on one image
par(mfrow=c(1,4))
for(i in 1:4) {
    boxplot(x[,i], main=names(dataset)[i])
}
```


```{r}
# barplot for class breakdown
plot(y)
```

```{r fig.asp=1}
# scatter plot matrix
featurePlot(x=x, y=y, plot="ellipse")
```

```{r}
# box and whisker plots for each attribute
featurePlot(x=x, y=y, plot="box")
```

```{r}
# density plots for each attribute by class value
scales <- list(x=list(relation="free"), y=list(relation="free"))
featurePlot(x=x, y=y, plot="density", scales=scales)
```

## Evaluate algorithms

### split and metrics

```{r}
# Run algorithms using 10-fold cross-validation
trainControl <- trainControl(method="cv", number=10)
metric <- "Accuracy"
```

### build models
```{r}
# LDA
set.seed(7)
fit.lda <- train(Species~., data=dataset, method = "lda", 
                 metric=metric, trControl=trainControl)
# CART
set.seed(7)
fit.cart <- train(Species~., data=dataset, method = "rpart", 
                  metric=metric, trControl=trainControl)
# KNN
set.seed(7)
fit.knn <- train(Species~., data=dataset, method = "knn", 
                 metric=metric, trControl=trainControl)
# SVM
set.seed(7)
fit.svm <- train(Species~., data=dataset, method = "svmRadial", 
                 metric=metric, trControl=trainControl)
# Random Forest
set.seed(7)
fit.rf <- train(Species~., data=dataset, method = "rf", 
                metric=metric, trControl=trainControl)
```

### compare
```{r}
#summarize accuracy of models
results <- resamples(list(lda  = fit.lda, 
                          cart = fit.cart, 
                          knn  = fit.knn, 
                          svm  = fit.svm, 
                          rf   = fit.rf))
summary(results)
```

```{r}
# compare accuracy of models
dotplot(results)
```

```{r}
# summarize Best Model
print(fit.lda)
```

## Make predictions

```{r}
# estimate skill of LDA on the validation dataset
predictions <- predict(fit.lda, validation)
confusionMatrix(predictions, validation$Species)
```

