# Building a Naive Bayes Classifier 

* Datasets: `iris.csv`
* Algorithms:
  * Naive Bayes


https://www.machinelearningplus.com/predictive-modeling/how-naive-bayes-algorithm-works-with-example-and-full-code/

## 8. Building a Naive Bayes Classifier in R

Understanding Naive Bayes was the (slightly) tricky part. Implementing it is fairly straightforward.

In R, Naive Bayes classifier is implemented in packages such as `e1071`, `klaR` and `bnlearn`. In Python, it is implemented in `scikit-learn`.

For sake of demonstration, let’s use the standard iris dataset to predict the Species of flower using 4 different features: Sepal.Length, Sepal.Width, Petal.Length, Petal.Width

```{r}
# Import Data
training <- read.csv('https://raw.githubusercontent.com/selva86/datasets/master/iris_train.csv')
test <- read.csv('https://raw.githubusercontent.com/selva86/datasets/master/iris_test.csv')
```

The training data is now contained in training and test data in test dataframe. Lets load the klaR package and build the naive bayes model.

```{r}
# Using klaR for Naive Bayes
library(klaR)
nb_mod <- NaiveBayes(Species ~ ., data=training)
pred <- predict(nb_mod, test)
```

Lets see the confusion matrix.

```{r}
# Confusion Matrix
tab <- table(pred$class, test$Species)
caret::confusionMatrix(tab)  
```

```{r}
# Plot density of each feature using nb_mod
opar = par(mfrow=c(2, 2), mar=c(4,0,0,0))
plot(nb_mod, main="")  
par(opar)
```

```{r}
# Plot the Confusion Matrix
library(ggplot2)
test$pred <- pred$class
ggplot(test, aes(Species, pred, color = Species)) +
  geom_jitter(width = 0.2, height = 0.1, size=2) +
  labs(title="Confusion Matrix", 
       subtitle="Predicted vs. Observed from Iris dataset", 
       y="Predicted", 
       x="Truth",
       caption="machinelearningplus.com")
```


