# Predicting the type of glass

* Datasets: `Glass`
* Algorithms:
    * SVM
    * Partitioning Tree

https://cran.r-project.org/web/packages/e1071/vignettes/svmdoc.pdf

In this example, we use the glass data from the UCI Repository of Machine
Learning Databases for classification. The task is to predict the type of a glass
on basis of its chemical analysis. We start by splitting the data into a train and
test set:

```{r}
library(caret)
library(e1071)
library(rpart)

data(Glass, package="mlbench")
str(Glass)
```

```{r}
## split data into a train and test set
index <- 1:nrow(Glass)
testindex <- sample(index, trunc(length(index)/3))
testset  <- Glass[testindex,]
trainset <- Glass[-testindex,]
```

Both for the SVM and the partitioning tree (via `rpart()`), we fit the model and
try to predict the test set values:

```{r}
## svm
svm.model <- svm(Type ~ ., data = trainset, cost = 100, gamma = 1)
svm.pred  <- predict(svm.model, testset[,-10])
```

(The dependent variable, Type, has column number 10. cost is a general penalizing
parameter for C-classification and `gamma` is the radial basis function-specific
`kernel` parameter.)

```{r}
## rpart
rpart.model <- rpart(Type ~ ., data = trainset)
rpart.pred <- predict(rpart.model, testset[,-10], type = "class")
```

A cross-tabulation of the true versus the predicted values yields:

```{r}
## compute svm confusion matrix
table(pred = svm.pred, true = testset[,10])
```


```{r}
## compute rpart confusion matrix
table(pred = rpart.pred, true = testset[,10])
```

### Comparison test sets

```{r}
confusionMatrix(svm.pred, testset$Type)
```

```{r}
confusionMatrix(rpart.pred, testset$Type)
```

### Comparison with resamples
Finally, we compare the performance of the two methods by computing the
respective accuracy rates and the kappa indices (as computed by `classAgreement()`
also contained in package `e1071`). In Table 1, we summarize the results
of 10 replications—Support Vector Machines show better results.

```{r}
set.seed(1234567)

# SVM
fit.svm <- train(Type ~., data = trainset, 
                 method = "svmRadial")

# Random Forest
fit.rpart <- train(Type ~., data = trainset, 
                method="rpart")

# collect resamples
results <- resamples(list(svm = fit.svm, 
                          rpart  = fit.rpart))

summary(results)
```


