# Applying Naive-Bayes on the Titanic case

* Datasets: `Titanic`
* Algorithms:
    * Naive Bayes
    


The Titanic dataset in R is a table for about 2200 passengers summarised according to four factors – economic status ranging from 1st class, 2nd class, 3rd class and crew; gender which is either male or female; Age category which is either Child or Adult and whether the type of passenger survived. For each combination of Age, Gender, Class and Survived status, the table gives the number of passengers who fall into the combination. We will use the Naive Bayes Technique to classify such passengers and check how well it performs. 

```{r}
#Getting started with Naive Bayes
#Install the package
#install.packages(“e1071”)
#Loading the library
library(e1071)

#Next load the Titanic dataset
data("Titanic")
#Save into a data frame and view it
Titanic_df = as.data.frame(Titanic)
```

We see that there are 32 observations which represent all possible combinations of Class, Sex, Age and Survived with their frequency. Since it is summarised, this table is not suitable for modelling purposes. We need to expand the table into individual rows. Let’s create a repeating sequence of rows based on the frequencies in the table

```{r}
#Creating data from table
repeating_sequence=rep.int(seq_len(nrow(Titanic_df)), Titanic_df$Freq) #This will repeat each combination equal to the frequency of each combination

# Create the dataset by row repetition created
Titanic_dataset=Titanic_df[repeating_sequence,]

# We no longer need the frequency, drop the feature
Titanic_dataset$Freq=NULL
```

The data is now ready for Naive Bayes to process. Let’s fit the model

```{r}
# Fitting the Naive Bayes model
Naive_Bayes_Model=naiveBayes(Survived ~., data=Titanic_dataset)

# What does the model say? Print the model summary
Naive_Bayes_Model
```

The model creates the conditional probability for each feature separately. We also have the a-priori probabilities which indicates the distribution of our data. Let’s calculate how we perform on the data.

```{r}
# Prediction on the dataset
NB_Predictions=predict(Naive_Bayes_Model,Titanic_dataset)
# Confusion matrix to check accuracy
table(NB_Predictions,Titanic_dataset$Survived)
```

We have the results! We are able to classify 1364 out of 1490 “No” cases correctly and 349 out of 711 “Yes” cases correctly. This means the ability of Naive Bayes algorithm to predict “No” cases is about 91.5% but it falls down to only 49% of the “Yes” cases resulting in an overall accuracy of 77.8%

## Can we Do any Better?
Naive Bayes is a parametric algorithm which implies that you cannot perform differently in different runs as long as the data remains the same. We will, however, learn another implementation of Naive Bayes algorithm using the ‘mlr’ package. Assuming the same session is going on for the readers, I will install and load the package and start fitting a model

```{r}
# Getting started with Naive Bayes in mlr
# install.packages(“mlr”)
# Loading the library
library(mlr)
```

The mlr package consists of a lot of models and works by creating tasks and learners which are then trained. Let’s create a classification task using the titanic dataset and fit a model with the naive bayes algorithm.

```{r}
# Create a classification task for learning on Titanic Dataset and specify the target feature
task = makeClassifTask(data = Titanic_dataset, target = "Survived")

# Initialize the Naive Bayes classifier
selected_model = makeLearner("classif.naiveBayes")

# Train the model
NB_mlr = train(selected_model, task)
```

The summary of the model which was printed in e3071 package is stored in learner model. Let’s print it and compare

```{r}
# Read the model learned  
NB_mlr$learner.model
```

The a-priori probabilities and the conditional probabilities for the model are similar to the one calculated by e3071 package as was expected. This means that our predictions will also be the same.

```{r}
# Predict on the dataset without passing the target feature
predictions_mlr = as.data.frame(predict(NB_mlr, newdata = Titanic_dataset[,1:3]))

## Confusion matrix to check accuracy
table(predictions_mlr[,1],Titanic_dataset$Survived)
```

As we see, the predictions are exactly same. The only way to improve is to have more features or more data. Perhaps, if we have more features such as the exact age, size of family, number of parents in the ship and siblings then we may arrive at a better model using Naive Bayes. In essence, Naive Bayes has an advantage of a strong foundation build and is very robust. I know of the ‘caret’ package which also consists of Naive Bayes function but it will also give us the same predictions and probability.


