# (PART) Linear Regression {.unnumbered}

# Linear Regression with ISLR

Dataset: `Advertising.csv`

Videos, slides:

* https://www.r-bloggers.com/in-depth-introduction-to-machine-learning-in-15-hours-of-expert-videos/

Data:

* http://www-bcf.usc.edu/~gareth/ISL/Advertising.csv

code:

* http://subasish.github.io/pages/ISLwithR/
* http://math480-s15-zarringhalam.wikispaces.umb.edu/R+Code
* https://github.com/yahwes/ISLR
* https://www.tau.ac.il/~saharon/IntroStatLearn.html
* https://www.waxworksmath.com/Authors/G_M/James/WWW/chapter_3.html
* https://github.com/asadoughi/stat-learning

plots:

* https://onlinecourses.science.psu.edu/stat857/node/28/

```{r}
library(readr)

advertising <- read_csv(file.path(data_raw_dir, "Advertising.csv"))
advertising
```

The Advertising data set. The plot displays sales, in thousands
of units, as a function of TV, radio, and newspaper budgets, in thousands of
dollars, for 200 diﬀerent markets. 

```{r}
par(mfrow=c(1,3))
plot(advertising$TV, advertising$sales, xlab = "TV", ylab = "Sales", col = "red")
plot(advertising$radio, advertising$sales, xlab="Radio", ylab="Sales", col="red")
plot(advertising$radio, advertising$newspaper, xlab="Newspaper", 
     ylab="Sales", col="red")
```

In each plot we show the simple least squares
ﬁt of sales to that variable, as described in Chapter 3. In other words, each blue
line represents a simple model that can be used to predict sales using TV, radio,
and newspaper, respectively.

```{r}
par(mfrow=c(1,3))
tv_model <- lm(sales ~ TV, data = advertising)
radio_model <- lm(sales ~ radio, data = advertising)
newspaper_model <- lm(sales ~ newspaper, data = advertising)

plot(advertising$TV, advertising$sales, xlab = "TV", ylab = "Sales", col = "red")
abline(tv_model, col = "blue")
plot(advertising$radio, advertising$sales, xlab="Radio", ylab="Sales", col="red")
abline(radio_model)
plot(advertising$newspaper, advertising$sales, xlab="Newspaper", 
     ylab="Sales", col="red")
abline(newspaper_model)
```

Recall the Advertising data from Chapter 2. Figure 2.1 displays sales
(in thousands of units) for a particular product as a function of advertis-
ing budgets (in thousands of dollars) for TV, radio, and newspaper media.
Suppose that in our role as statistical consultants we are asked to suggest,
on the basis of this data, a marketing plan for next year that will result in
high product sales. What information would be useful in order to provide
such a recommendation? Here are a few important questions that we might
seek to address:

1.  Is there a relationship between advertising budget and sales?

2. How strong is the relationship between advertising budget and sales?

3. Which media contribute to sales?

4. How accurately can we estimate the eﬀect of each medium on sales?


 For the Advertising data, the least squares fit for the regression
of sales onto TV is shown. The fit is found by minimizing the sum of squared
errors. Each grey line segment represents an error, and the fit makes a compro-
mise by averaging their squares. In this case a linear fit captures the essence of
the relationship, although it is somewhat deficient in the left of the plot.

```{r}
tv_model <- lm(sales ~ TV, data = advertising)
plot(advertising$TV, advertising$sales, xlab = "TV", ylab = "Sales", 
     col = "red", pch=16)
abline(tv_model, col = "blue", lwd=2)
segments(advertising$TV, advertising$sales, advertising$TV, predict(tv_model), 
         col = "gray")
```

```{r}
smry <- summary(tv_model)
smry
```

```{r}
library(lattice)

minRss <- sqrt(abs(min(smry$residuals))) * sign(min(smry$residuals))
maxRss <- sqrt(max(smry$residuals))

twovar <- function(x, y) { 
  x^2 + y^2 }

mat <- outer( seq(minRss, maxRss, length = 100),  
                seq(minRss, maxRss, length = 100), 
                Vectorize( function(x,y) twovar(x, y) ) )



contourplot(mat, at = c(1,2,3))
```

```{r}
tv_model
```

```{r}
tv.lm <- lm(sales ~ poly(sales, TV, degree=2), data = advertising)
# contour(tv.lm, sales ~ TV)
```

```{r}
library(rsm)
mpg.lm <- lm(mpg ~ poly(hp, disp, degree = 3), data = mtcars)
contour(mpg.lm, hp ~ disp)
```

```{r}
x <- -6:16
op <- par(mfrow = c(2, 2))
contour(outer(x, x), method = "flattest", vfont = c("sans serif", "plain"))
```


