# Evaluation of three linear regression models

* Dataset: `iris.csv`
* Algorithms: 
  * Simple Linear Regression
  * Multiple Regression
  * Neural Networks

```{r echo=FALSE}
# load(file.path(rprojroot::find_rstudio_root_file(), "workspace.RData"))
#source(file.path(rprojroot::find_rstudio_root_file(), 'R', 'set_folders.R'))
#.libPaths(c(normalizePath(sbox_path), normalizePath(lib_path), .libPaths()))
```

## Introduction

https://www.matthewrenze.com/workshops/practical-machine-learning-with-r/lab-3a-regression.html


## Explore the Data
1. Load Iris data
2. Plot scatterplot
3. Plot correlogram


```{r}
data(iris)
```

```{r}
write.csv(iris, file.path(data_raw_dir, "iris.csv"))
```


Create scatterplot matrix

```{r}
plot(iris[1:4])
```


```{r}
library(corrgram)
corrgram(iris[1:4])
```


```{r}
cor(iris[1:4])

```


```{r}
cor(
  x = iris$Petal.Length, 
  y = iris$Petal.Width)
```


```{r}
plot(
  x = iris$Petal.Length, 
  y = iris$Petal.Width,
  xlim = c(0.25, 7),
  ylim = c(0.25, 2.5))
```

## Create Training and Test Sets

```{r}
set.seed(42)
```


```{r}
indexes <- sample(
  x = 1:150, 
  size = 100)
```


```{r}
train <- iris[indexes, ]
test <- iris[-indexes, ]
```


## Predict with Simple Linear Regression

```{r}
simpleModel <- lm(
  formula = Petal.Width ~ Petal.Length,
  data = train)
```


```{r}
plot(
  x = iris$Petal.Length, 
  y = iris$Petal.Width,
  xlim = c(0.25, 7),
  ylim = c(0.25, 2.5))
  
lines(
  x = train$Petal.Length,
  y = simpleModel$fitted, 
  col = "red",
  lwd = 3)
```


```{r}
summary(simpleModel)
```


```{r}
simplePredictions <- predict(
  object = simpleModel,
  newdata = test)
```


```{r}
plot(
  x = iris$Petal.Length, 
  y = iris$Petal.Width,
  xlim = c(0.25, 7),
  ylim = c(0.25, 2.5))
  
points(
  x = test$Petal.Length,
  y = simplePredictions,
  col = "blue",
  pch = 4,
  lwd = 2)

points(
  x = test$Petal.Length,
  y = test$Petal.Width,
  col = "red",
  pch = 16)
```



```{r}
simpleRMSE <- sqrt(mean((test$Petal.Width - simplePredictions)^2))
print(simpleRMSE)
```

## Predict with Multiple Regression

```{r}
multipleModel <- lm(
  formula = Petal.Width ~ .,
  data = train)
```

```{r}
summary(multipleModel)
```

```{r}
multiplePredictions <- predict(
  object = multipleModel,
  newdata = test)
```

```{r}
plot(
  x = iris$Petal.Length, 
  y = iris$Petal.Width,
  xlim = c(0.25, 7),
  ylim = c(0.25, 2.5))
  
points(
  x = test$Petal.Length,
  y = multiplePredictions,
  col = "blue",
  pch = 4,
  lwd = 2)

points(
  x = test$Petal.Length,
  y = test$Petal.Width,
  col = "red",
  pch = 16)
```


```{r}
multipleRMSE <- sqrt(mean((test$Petal.Width - multiplePredictions)^2))
print(multipleRMSE)
```


## 5. Predict with Neural Network Regression

```{r}
normalize <- function(x) {
  (x - min(x)) / (max(x) - min(x)) - 0.5
}
```

```{r}
denormalize <- function(x, y) {
  ((x + 0.5) * (max(y) - min(y))) + min(y)
}
```


```{r}
scaledIris <- data.frame(
  Sepal.Length = normalize(iris$Sepal.Length),
  Sepal.Width = normalize(iris$Sepal.Width),
  Petal.Length = normalize(iris$Petal.Length),
  Petal.Width = normalize(iris$Petal.Width),
  Species = iris$Species)
```


```{r}
scaledTrain <- scaledIris[indexes, ]
scaledTest <- scaledIris[-indexes, ]
```

```{r run-nnet}
library(nnet)

neuralRegressor <- nnet(
  formula = Petal.Width ~ .,
  data = scaledTrain,
  linout = TRUE,
  skip = TRUE,
  size = 4,
  decay = 0.0001,
  maxit = 500)
```


```{r}
library(NeuralNetTools)

plotnet(neuralRegressor)
```


```{r}
scaledPredictions <- predict(
  object = neuralRegressor, 
  newdata = scaledTest)
```


```{r}
neuralPredictions <- denormalize(
  x = scaledPredictions, 
  y = iris$Petal.Width)
```


```{r}
plot(
  x = iris$Petal.Length, 
  y = iris$Petal.Width,
  xlim = c(0.25, 7),
  ylim = c(0.25, 2.5))
  
points(
  x = test$Petal.Length,
  y = neuralPredictions,
  col = "blue",
  pch = 4,
  lwd = 2)

points(
  x = test$Petal.Length,
  y = test$Petal.Width,
  col = "red",
  pch = 16)
```

```{r}
neuralRMSE <- sqrt(mean((test$Petal.Width - neuralPredictions)^2))
print(neuralRMSE)
```

## 6. Evaluate all the regression Models

```{r}
print(simpleRMSE)
print(multipleRMSE)
print(neuralRMSE)
```



