# Temperature modeling using nested dataframes

Dataset: `temperature.csv`

## Introduction

[http://ijlyttle.github.io/isugg_purrr/presentation.html\#(1)](http://ijlyttle.github.io/isugg_purrr/presentation.html#(1)){.uri}

### Packages to run this presentation

```{r packages, message=FALSE, warning=FALSE}
library("readr")
library("tibble")
library("dplyr")
library("tidyr")
library("stringr")
library("ggplot2")
library("purrr")
library("broom")
```

### Motivation

As you know, purrr is a recent package from Hadley Wickham, focused on lists and functional programming, like dplyr is focused on data-frames.

I figure a good way to learn a new package is to try to solve a problem, so we have a dataset:

-   you can [view](https://github.com/ijlyttle/isugg_purrr/blob/gh-pages/temperature.csv) or [download](http://ijlyttle.github.io/isugg_purrr/temperature.csv)

-   you can download the [source](http://ijlyttle.github.io/isugg_purrr/presentation.Rmd) of this presentation

-   these are three temperatures recorded simultaneously in a piece of electronics

-   it will be very valuable to be able to characterize the transient temperature for each sensor

-   we want to apply the same set of models across all three sensors

-   it will be easier to show using pictures

## Prepare the data

### Let's get the data into shape

Using the readr package

```{r load}
temperature_wide <- 
  read_csv(file.path(data_raw_dir, "temperature.csv")) %>%
  print()
```

### Is `temperature_wide` "tidy"?

```{r is_tidy, echo=FALSE}
print(temperature_wide)
```

Why or why not?

### Tidy data

1.  Each column is a variable
2.  Each row is an observation
3.  Each cell is a value

(<http://www.jstatsoft.org/v59/i10/paper>)

My personal observation is that "tidy" can depend on the context, on what you want to do with the data.

### Let's get this into a tidy form

```{r}
temperature_tall <-
  temperature_wide %>%
  gather(key = "id_sensor", value = "temperature", starts_with("temp")) %>%
  mutate(id_sensor = str_replace(id_sensor, "temperature_", "")) %>%
  print()
```

### Now, it's easier to visualize

```{r}
temperature_tall %>%
  ggplot(aes(x = instant, y = temperature, color = id_sensor)) +
  geom_line()
```

### Calculate delta time ($\Delta t$) and delta temperature ($\Delta T$)

**`delta_time`** $\Delta t$

change in time since event started, s

**`delta_temperature`**: $\Delta T$

change in temperature since event started, °C

```{r}
delta <- 
  temperature_tall %>%
  arrange(id_sensor, instant) %>%
  group_by(id_sensor) %>%
  mutate(
    delta_time = as.numeric(instant) - as.numeric(instant[[1]]),
    delta_temperature = temperature - temperature[[1]]
  ) %>%
  select(id_sensor, delta_time, delta_temperature)
```

### Let's have a look

```{r}
# plot delta time vs delta temperature, by sensor
delta %>%
  ggplot(aes(x = delta_time, y = delta_temperature, color = id_sensor)) +
  geom_line()  
```

## Define the models

We want to see how three different curve-fits might perform on these three data-sets:

#### Newtonian cooling

$$\Delta T = \Delta {T_0} * (1 - e^{-\frac{\delta t}{\tau_0}})$$

### Semi-infinite solid

$$\Delta T = \Delta T_0 * erfc(\sqrt{\frac{\tau_0}{\delta t}}))$$

### Semi-infinite solid with convection

$$\Delta T = \Delta T_0 * \big [ \operatorname erfc(\sqrt{\frac{\tau_0}{\delta t}}) - e^ {Bi_0 + (\frac {Bi_0}{2})^2 \frac {\delta t}{\tau_0}} * \operatorname erfc (\sqrt \frac{\tau_0}{\delta t} + \frac {Bi_0}{2} * \sqrt \frac{\delta t }{\tau_0} \big]$$

### `erf` and `erfc` functions

```{r}
# reference: http://stackoverflow.com/questions/29067916/r-error-function-erfz
# (see Abramowitz and Stegun 29.2.29)
erf <- function(x) 2 * pnorm(x * sqrt(2)) - 1
erfc <- function(x) 2 * pnorm(x * sqrt(2), lower = FALSE)
```

### Newton cooling equation

```{r}
newton_cooling <- function(x) {
  nls(
    delta_temperature ~ delta_temperature_0 * (1 - exp(-delta_time/tau_0)),
    start = list(delta_temperature_0 = -10, tau_0 = 50),
    data = x
  )
}
```

### Temperature models: simple and convection

```{r}
semi_infinite_simple <- function(x) {
  nls(
    delta_temperature ~ delta_temperature_0 * erfc(sqrt(tau_0 / delta_time)),
    start = list(delta_temperature_0 = -10, tau_0 = 50),
    data = x
  )    
}

semi_infinite_convection <- function(x){
  nls(
    delta_temperature ~
      delta_temperature_0 * (
        erfc(sqrt(tau_0 / delta_time)) -
        exp(Bi_0 + (Bi_0/2)^2 * delta_time / tau_0) *
          erfc(sqrt(tau_0 / delta_time) + 
        (Bi_0/2) * sqrt(delta_time / tau_0))
      ),
    start = list(delta_temperature_0 = -5, tau_0 = 50, Bi_0 = 1.e6),
    data = x
  )
}
```

## Test modeling on one dataset

### Before going into `purrr`

Before doing anything, we want to show that we can do something with one dataset and one model-function:

```{r}
# only one sensor; it is a test
tmp_data <- delta %>% filter(id_sensor == "a")

tmp_model <- newton_cooling(tmp_data)

summary(tmp_model)
```

### Look at predictions

```{r}
# apply prediction and make it tidy
tmp_pred <- 
  tmp_data %>%
  mutate(modeled = predict(tmp_model, data = .)) %>%
  select(id_sensor, delta_time, measured = delta_temperature, modeled) %>%
  gather("type", "delta_temperature", measured:modeled) %>%
  print()
```

### Plot Newton model

```{r}
tmp_pred %>%
  ggplot(aes(x = delta_time, y = delta_temperature, linetype = type)) +
  geom_line() +
  labs(title = "Newton temperature model", subtitle = "One sensor: a")
```

### "Regular" data-frame (deltas)

```{r}
print(delta)
```

Each column of the dataframe is a vector - in this case, a character vector and two doubles

## Making a nested dataframe

### How to make a weird data-frame

Here's where the fun starts - a column of a data-frame can be a list.

-   use `tidyr::nest()` to makes a column `data`, which is a list of data-frames

-   this seems like a stronger expression of the `dplyr::group_by()` idea

```{r}
# nest delta_time and delta_temperature variables
delta_nested <- 
  delta %>%
  nest(-id_sensor) %>%
  print()
```

### Map dataframes to a modeling function (Newton)

-   `map()` is like `lapply()`

-   `map()` returns a list-column (it keeps the weirdness)

```{r}
model_nested <-
  delta_nested %>%
  mutate(model = map(data, newton_cooling)) %>%
  print()
```

> We get an additional list-column `model`.

### We can use `map2()` to make the predictions

-   `map2()` is like `mapply()`

-   designed to map two colunms (`model`, `data`) to a function `predict()`

```{r}
predict_nested <-
  model_nested %>%
  mutate(pred = map2(model, data, predict)) %>%
  print()
```

> Another list-column `pred` for the prediction results.

### We need to get out of the weirdness

-   use `unnest()` to get back to a regular data-frame

```{r}
predict_unnested <- 
  predict_nested %>%
  unnest(data, pred) %>% 
  print()
```

### We can wrangle the predictions

-   get into a form that makes it easier to plot

```{r}
predict_tall <- 
  predict_unnested %>%
  rename(modeled = pred, measured = delta_temperature) %>%
  gather("type", "delta_temperature", modeled, measured) %>%
  print()
```

### We can visualize the predictions

```{r}
predict_tall %>%
  ggplot(aes(x = delta_time, y = delta_temperature)) +
  geom_line(aes(color = id_sensor, linetype = type)) +
  labs(title = "Newton temperature modeling", 
       subtitle = "Three sensors: a, b, c")
```

## Apply multiple models on a nested structure

### Step 1: Selection of models

Make a list of functions to model:

```{r}
list_model <-
  list(
    newton_cooling = newton_cooling,
    semi_infinite_simple = semi_infinite_simple,
    semi_infinite_convection = semi_infinite_convection
  )
```

### Step 2: write a function to define the "inner" loop

```{r}
# add additional variable with the model name

fn_model <- function(.model, df) {
  # one parameter for the model in the list, the second for the data
  # safer to avoid non-standard evaluation
  # df %>% mutate(model = map(data, .model)) 
  
  df$model <- map(df$data, possibly(.model, NULL))
  df
}
```

-   for a given model-function and a given (weird) data-frame, return a modified version of that data-frame with a column `model`, which is the model-function applied to each element of the data-frame's `data` column (which is itself a list of data-frames)

-   the purrr functions `safely()` and `possibly()` are **very** interesting. I think they could be useful outside of purrr as a friendlier way to do error-handling.

### Step 3: Use `map_df()` to define the "outer" loop

```{r}
# this dataframe will be the second input of fn_model
delta_nested %>% 
  print()
```

```{r}
# fn_model is receiving two inputs: one from list_model and from delta_nested
model_nested_new <-
  list_model %>%
  map_df(fn_model, delta_nested, .id = "id_model") %>%
  print()
```

-   for each element of a list of model-functions, run the inner-loop function, and row-bind the results into a data-frame

-   we want to discard the rows where the model failed

-   we also want to investigate why they failed, but that's a different talk

### Step 4: Use `map()` to identify the null models

```{r}
model_nested_new <-
  list_model %>%
  map_df(fn_model, delta_nested, .id = "id_model") %>%
  mutate(is_null = map(model, is.null)) %>%
  print()
```

-   using `map(model, is.null)` returns a list column
-   to use `filter()`, we have to escape the weirdness

### Step 5: `map_lgl()` to identify nulls and get out of the weirdness

```{r}
model_nested_new <-
  list_model %>%
  map_df(fn_model, delta_nested, .id = "id_model") %>%
  mutate(is_null = map_lgl(model, is.null)) %>%
  print()
```

-   using `map_lgl(model, is.null)` returns a vector column

### Step 6: `filter()` nulls and `select()` variables to clean up

```{r}
model_nested_new <-
  list_model %>%
  map_df(fn_model, delta_nested, .id = "id_model") %>%
  mutate(is_null = map_lgl(model, is.null)) %>%
  filter(!is_null) %>%
  select(-is_null) %>%
  print()
```

### Step 7: Calculate predictions on nested dataframe

```{r}
predict_nested <- 
  model_nested_new %>%
  mutate(pred = map2(model, data, predict)) %>%
  print()
```

### `unnest()`, make it tall and tidy

```{r}
predict_tall <-
  predict_nested %>%
  unnest(data, pred) %>% 
  rename(modeled = pred, measured = delta_temperature) %>%
  gather("type", "delta_temperature", modeled, measured) %>%
  print()
```

### Visualize the predictions

```{r}
predict_tall %>%
  ggplot(aes(x = delta_time, y = delta_temperature)) +
  geom_line(aes(color = id_sensor, linetype = type)) +
  facet_grid(id_model ~ .) +
  labs(title = "Newton and Semi-infinite temperature modeling", 
       subtitle = "Three sensors: a, b, c")
```

### Let's get the residuals

```{r}
resid <-
  model_nested_new %>%
  mutate(resid = map(model, resid)) %>%
  unnest(data, resid) %>%
  print()
```

### And visualize them

```{r}
resid %>%
  ggplot(aes(x = delta_time, y = resid)) +
  geom_line(aes(color = id_sensor)) +
  facet_grid(id_model ~ .) +
  labs(title = "Residuals for Newton and Semi-infinite models")
```

## Using broom package to look at model-statistics

We will use a previous defined dataframe with the model and data:

```{r}
model_nested_new %>% 
  print()
```

The `tidy()` function extracts statistics from a model.

```{r}
# apply over model_nested_new but only three variables
model_parameters <- 
  model_nested_new %>%
  select(id_model, id_sensor, model) %>%
  mutate(tidy = map(model, tidy)) %>%
  select(-model) %>%
  unnest() %>%
  print()
```

### Get a sense of the coefficients

```{r}
model_summary <-
  model_parameters %>%
  select(id_model, id_sensor, term, estimate) %>%
  spread(key = "term", value = "estimate") %>%
  print()
```

### Summary

-   this is just a smalll part of purrr

-   there seem to be parallels between `tidyr::nest()/purrr::map()` and `dplyr::group_by()/dplyr::do()`

    -   to my mind, the purrr framework is more understandable
    -   update tweet from [Hadley](https://twitter.com/hadleywickham/status/719542847045636096)

References from Hadley:

-   [purrr 0.1.0 announcement](http://blog.rstudio.org/2015/09/29/purrr-0-1-0/)
-   [purrr 0.2.0 announcement](http://blog.rstudio.org/2016/01/06/purrr-0-2-0/)
-   [chapter from Garrett Grolemund and Hadley's forthcoming book](http://r4ds.had.co.nz/iteration.html)
