# Visualization of neural networks

https://beckmw.wordpress.com/tag/neuralnet/

In my last post I said I wasn’t going to write anymore about neural networks (i.e., multilayer feedforward perceptron, supervised ANN, etc.). That was a lie. I’ve received several requests to update the neural network plotting function described in the original post. As previously explained, R does not provide a lot of options for visualizing neural networks. The only option I know of is a plotting method for objects from the neuralnet package. This may be my opinion, but I think this plot leaves much to be desired (see below). Also, no plotting methods exist for neural networks created in other packages, i.e., nnet and RSNNS. These packages are the only ones listed on the CRAN task view, so I’ve updated my original plotting function to work with all three. Additionally, I’ve added a new option for plotting a raw weight vector to allow use with neural networks created elsewhere. This blog describes these changes, as well as some new arguments added to the original function.

As usual, I’ll simulate some data to use for creating the neural networks. The dataset contains eight input variables and two output variables. The final dataset is a data frame with all variables, as well as separate data frames for the input and output variables. I’ve retained separate datasets based on the syntax for each package.

```{r}
library(clusterGeneration)
library(tictoc)
 
seed.val<- 12345
set.seed(seed.val)
 
num.vars<-8
num.obs<-1000
 
# input variables
cov.mat <-genPositiveDefMat(num.vars,covMethod=c("unifcorrmat"))$Sigma
rand.vars <-mvrnorm(num.obs,rep(0,num.vars),Sigma=cov.mat)
 
# output variables
parms <-runif(num.vars,-10,10)
y1 <- rand.vars %*% matrix(parms) + rnorm(num.obs,sd=20)
parms2 <- runif(num.vars,-10,10)
y2 <- rand.vars %*% matrix(parms2) + rnorm(num.obs,sd=20)
 
# final datasets
rand.vars <- data.frame(rand.vars)
resp <- data.frame(y1,y2)
names(resp) <- c('Y1','Y2')
dat.in <- data.frame(resp, rand.vars)
```

```{r}
dplyr::glimpse(dat.in)
```


The various neural network packages are used to create separate models for plotting.

```{r run-nnet}
# first model with nnet
#nnet function from nnet package
library(nnet)
set.seed(seed.val)
tic()
mod1 <- nnet(rand.vars, resp, data = dat.in, size = 10, linout = T)
toc()
```

```{r}
# nn <- neuralnet(form.in,
#                 data = dat.sc,
#                 # hidden = c(13, 10, 3),
#                 hidden = c(5),
#                 act.fct = "tanh",
#                 linear.output = FALSE,
#                 lifesign = "minimal")
```

```{r run-neuralnet}
# 2nd model with neuralnet
# neuralnet function from neuralnet package, notice use of only one response
library(neuralnet)

softplus <- function(x) log(1 + exp(x))
sigmoid  <- function(x) log(1 + exp(-x))

dat.sc <- scale(dat.in)
form.in <- as.formula('Y1 ~ X1+X2+X3+X4+X5+X6+X7+X8')
set.seed(seed.val)
tic()
mod2 <- neuralnet(form.in, data = dat.sc, hidden = 10, lifesign = "minimal",
                  linear.output = FALSE,
                  act.fct = "tanh")
toc()
```


```{r run-rsnns}
# third model with RSNNS
# mlp function from RSNNS package
library(RSNNS)
set.seed(seed.val)
tic()
mod3 <- mlp(rand.vars, resp, size = 10, linOut = T)
toc()
```

I’ve noticed some differences between the functions that could lead to some confusion. For simplicity, the above code represents my interpretation of the most direct way to create a neural network in each package. Be very aware that direct comparison of results is not advised given that the default arguments differ between the packages. A few key differences are as follows, although many others should be noted. First, the functions differ in the methods for passing the primary input variables. 

The `nnet` function can take separate (or combined) x and y inputs as data frames or as a formula, the `neuralnet` function can only use a formula as input, and the mlp function can only take a data frame as combined or separate variables as input. As far as I know, the neuralnet function is not capable of modelling multiple response variables, unless the response is a categorical variable that uses one node for each outcome. Additionally, the default output for the neuralnet function is linear, whereas the opposite is true for the other two functions.

Specifics aside, here’s how to use the updated plot function. Note that the same syntax is used to plot each model


```{r}
# import the function from Github
library(devtools)
source_url('https://gist.githubusercontent.com/fawda123/7471137/raw/466c1474d0a505ff044412703516c34f1a4684a5/nnet_plot_update.r')
 
# plot each model
plot.nnet(mod1)
plot.nnet(mod2) 
plot.nnet(mod3)
```

The plotting function can also now be used with an arbitrary weight vector, rather than a specific model object. The struct argument must also be included if this option is used. I thought the easiest way to use the plotting function with your own weights was to have the input weights as a numeric vector, including bias layers. I’ve shown how this can be done using the weights directly from mod1 for simplicity.

```{r}
wts.in <- mod1$wts
struct <- mod1$n
plot.nnet(wts.in,struct=struct)
```

Note that `wts.in` is a numeric vector with length equal to the expected given the architecture (i.e., for 8 10 2 network, 100 connection weights plus 12 bias weights). The plot should look the same as the plot for the neural network from nnet.

The weights in the input vector need to be in a specific order for correct plotting. I realize this is not clear by looking directly at wt.in but this was the simplest approach I could think of. The weight vector shows the weights for each hidden node in sequence, starting with the bias input for each node, then the weights for each output node in sequence, starting with the bias input for each output node. Note that the bias layer has to be included even if the network was not created with biases. If this is the case, simply input a random number where the bias values should go and use the argument bias=F. I’ll show the correct order of the weights using an example with plot.nn from the neuralnet package since the weights are included directly on the plot.


If we pretend that the above figure wasn’t created in R, we would input the mod.in argument for the updated plotting function as follows. Also note that struct must be included if using this approach.

```{r}
mod.in<-c(13.12,1.49,0.16,-0.11,-0.19,-0.16,0.56,-0.52,0.81)
struct<-c(2,2,1) #two inputs, two hidden, one output 
plot.nnet(mod.in, struct=struct)
```

Note the comparability with the figure created using the neuralnet package. That is, larger weights have thicker lines and color indicates sign (+ black, – grey).

One of these days I’ll actually put these functions in a package. In the meantime, please let me know if any bugs are encountered.


## caret and plot NN
I’ve changed the function to work with neural networks created using the train function from the caret package. The link above is updated but you can also grab it here.

```{r run-nnet-training, results='hide'}
library(caret)
mod4 <- train(Y1 ~., method='nnet', data=dat.in, linout=T)
```

```{r}
plot.nnet(mod4,nid=T)
```


```{r}
fact<-factor(sample(c('a','b','c'),size=num.obs,replace=T))
form.in<-formula('cbind(Y2,Y1)~X1+X2+X3+fact')
mod5<-nnet(form.in,data=cbind(dat.in,fact),size=10,linout=T)
plot.nnet(mod5,nid=T)
```

## Multiple hidden layers
More updates… I’ve now modified the function to plot multiple hidden layers for networks created using the mlp function in the `RSNNS` package and neuralnet in the neuralnet package. As far as I know, these are the only neural network functions in R that can create multiple hidden layers. All others use a single hidden layer. I have not tested the plotting function using manual input for the weight vectors with multiple hidden layers. My guess is it won’t work but I can’t be bothered to change the function unless it’s specifically requested. The updated function can be grabbed here (all above links to the function have also been changed).

```{r run-rsnns-hidden_layers}
library(RSNNS)
 
# neural net with three hidden layers, 9, 11, and 8 nodes in each
tic()
mod <-mlp(rand.vars, resp, 
          size = c(9,11,8), 
          linOut = T)
toc()
par(mar=numeric(4),family='serif')
plot.nnet(mod)
```

## Binary predictors
Here’s an example using the `neuralnet` function with binary predictors and categorical outputs (credit to Tao Ma for the model code).

```{r}
library(neuralnet)
 
#response
AND<-c(rep(0,7),1)
OR<-c(0,rep(1,7))
 
# response with predictors
binary.data <- data.frame(expand.grid(c(0,1), c(0,1), c(0,1)), AND, OR)
 
#model
tic()
net <- neuralnet(AND+OR ~ Var1+Var2+Var3,
                 binary.data, hidden =c(6,12,8), 
                 rep = 10, 
                 err.fct="ce", 
                 linear.output=FALSE)
toc()
#plot ouput
par(mar=numeric(4),family='serif')
plot.nnet(net)
```

## color coding the input layer
The color vector argument (circle.col) for the nodes was changed to allow a separate color vector for the input layer. 

The following example shows how this can be done using relative importance of the input variables to color-code the first layer.

```{r}
# example showing use of separate colors for input layer
# color based on relative importance using 'gar.fun'
 
##
#create input data
seed.val<-3
set.seed(seed.val)
  
num.vars<-8
num.obs<-1000
  
#input variables
library(clusterGeneration)
cov.mat<-genPositiveDefMat(num.vars,covMethod=c("unifcorrmat"))$Sigma
rand.vars<-mvrnorm(num.obs,rep(0,num.vars),Sigma=cov.mat)
  
# output variables
parms<-runif(num.vars,-10,10)
y1<-rand.vars %*% matrix(parms) + rnorm(num.obs,sd=20)
 
# final datasets
rand.vars<-data.frame(rand.vars)
resp<-data.frame(y1)
names(resp)<-'Y1'
dat.in <- data.frame(resp,rand.vars)
 
##
# create model
library(nnet)
mod1 <- nnet(rand.vars,resp,data=dat.in,size=10,linout=T)
 
##
# relative importance function
library(devtools)
source_url('https://gist.github.com/fawda123/6206737/raw/2e1bc9cbc48d1a56d2a79dd1d33f414213f5f1b1/gar_fun.r')
 
# relative importance of input variables for Y1
rel.imp <- gar.fun('Y1',mod1,bar.plot=F)$rel.imp
 
#color vector based on relative importance of input values
cols<-colorRampPalette(c('green','red'))(num.vars)[rank(rel.imp)]
 
##
#plotting function
source_url('https://gist.githubusercontent.com/fawda123/7471137/raw/466c1474d0a505ff044412703516c34f1a4684a5/nnet_plot_update.r')
  
#plot model with new color vector
#separate colors for input vectors using a list for 'circle.col'
plot(mod1,circle.col=list(cols,'lightblue'))
```

