# QQ and PP Plots

https://homepage.divms.uiowa.edu/~luke/classes/STAT4580/qqpp.html

## QQ Plot
One way to assess how well a particular theoretical model describes a data distribution is to plot data quantiles against theoretical quantiles.

Base graphics provides `qqnorm`, lattice has `qqmath`, and ggplot2 has `geom_qq`.

The default theoretical distribution used in these is a standard normal, but, except for `qqnorm`, these allow you to specify an alternative.

For a large sample from the theoretical distribution the plot should be a straight line through the origin with slope 1:

```{r}
library(ggplot2)

n <- 10000
ggplot() + geom_qq(aes(sample = rnorm(n)))
```

If the plot is a straight line with a different slope or intercept, then the data distribution corresponds to a location-scale transformation of the theoretical distribution.

The slope is the scale and the intercept is the location:

```{r}
ggplot() +
    geom_qq(aes(sample = rnorm(n, 10, 4))) +
    geom_abline(intercept = 10, slope = 4,
                color = "red", size = 1.5, alpha = 0.8)

```

The QQ plot can be constructed directly as a scatterplot of the sorted sample $i = 1, \dots, n$ against quantiles for 

$$p_i = \frac{i}{n} - \frac{1}{2n}$$

```{r}
p <- (1 : n) / n - 0.5 / n
y <- rnorm(n, 10, 4)
ggplot() + geom_point(aes(x = qnorm(p), y = sort(y)))
```

## Some Examples
The histograms and density estimates for the duration variable in the `geyser` data set showed that the distribution is far from a normal distribution, and the normal QQ plot shows this as well:

```{r}
library(MASS)
ggplot(geyser) + geom_qq(aes(sample = duration))

```

Except for rounding the parent heights in the Galton data seemed not too fat from normally distributed:

```{r psych_galton, warning=FALSE, message=FALSE}
library(psych)
library(UsingR)

ggplot(galton) + geom_qq(aes(sample = parent))

```

Rounding interferes more with this visualization than with a histogram or a density plot.

Rounding is more visible with this visualization than with a histogram or a density plot.

Another Gatlton dataset available in the UsingR package with less rounding is father.son:

```{r usingr-father_son, warning=FALSE, message=FALSE}
library(UsingR)
ggplot(father.son) + geom_qq(aes(sample = fheight))

```

The middle seems to be fairly straight, but the ends are somewhat wiggly.

How can you calibrate your judgment?


## Calibrating the Variability
One approach is to use simulation, sometimes called a graphical bootstrap.

The `nboot` function will simulate R samples from a normal distribution that match a variable x on sample size, sample mean, and sample SD.

The result is returned in a dataframe suitable for plotting:

```{r graphical_boostrap}
nsim <- function(n, m = 0, s = 1) {
    z <- rnorm(n)
    m + s * ((z - mean(z)) / sd(z))
}

nboot <- function(x, R) {
    n <- length(x)
    m <- mean(x)
    s <- sd(x)
    do.call(rbind,
            lapply(1 : R,
                   function(i) {
                       xx <- sort(nsim(n, m, s))
                       p <- seq_along(x) / n - 0.5 / n
                       data.frame(x = xx, p = p, sim = i)
    }))
}
```

Plotting these as lines shows the variability in shapes we can expect when sampling from the theoretical normal distribution:

```{r}
gb <- nboot(father.son$fheight, 50)
tibble::as_tibble(gb)
```


```{r}
ggplot() +
    geom_line(aes(x = qnorm(p), y = x, group = sim),
              color = "gray", data = gb)
```

We can then insert this simulation behind our data to help calibrate the visualization:

```{r}
ggplot(father.son) +
    geom_line(aes(x = qnorm(p), y = x, group = sim),
              color = "gray", data = gb) +
    geom_qq(aes(sample = fheight))
```

## Scalability
For large sample sizes overplotting will occur:

```{r}
ggplot(diamonds) + geom_qq(aes(sample = price))

```

This can be alleviated by using a grid of quantiles:

```{r}
nq <- 100
p <- (1 : nq) / nq - 0.5 / nq
ggplot() + geom_point(aes(x = qnorm(p), y = quantile(diamonds$price, p)))

```

A more reasonable model might be an exponential distribution:

```{r}
ggplot() + geom_point(aes(x = qexp(p), y = quantile(diamonds$price, p)))

```


## Comparing Two Distributions
The QQ plot can also be used to compare two distributions based on a sample from each.

If the samples are the same size then this is just a plot of the ordered sample values against each other.

Choosing a fixed set of quantiles allows samples of unequal size to be compared.

Using a small set of quantiles we can compare the distributions of waiting times between eruptions of Old Faithful from the two different data sets we have looked at:

```{r compare_distributions}
nq <- 31  # user defined
nq <- min(length(geyser$waiting), length(faithful$waiting)) # or take the minimum
p <- (1 : nq) / nq - 0.5 / nq

wg <- geyser$waiting
wf <- faithful$waiting

ggplot() + geom_point(aes(x = quantile(wg, p), y = quantile(wf, p)))
```


## PP Plots
The PP plot for comparing a sample to a theoretical model plots the theoretical proportion less than or equal to each observed value against the actual proportion.

For a theoretical cumulative distribution function F this means plotting

$$F(x(i))∼pi$$

For the `fheight` variable in the `father.son` data:

```{r}
m <- mean(father.son$fheight)
s <- sd(father.son$fheight)
n <- nrow(father.son)
p <- (1 : n) / n - 0.5 / n
ggplot(father.son) + geom_point(aes(x = p, y = sort(pnorm(fheight, m, s))))
```

* The values on the vertical axis are the probability integral transform of the data for the theoretical distribution.

* If the data are a sample from the theoretical distribution then these transforms would be uniformly distributed on [0,1].

* The PP plot is a QQ plot of these transformed values against a uniform distribution.

* The PP plot goes through the points (0,0) and (1,1) and so is much less variable in the tails:


```{r}
pp <- ggplot() +
geom_line(aes(x = p, y = pnorm(x, m, s), group = sim),
          color = "gray", data = gb)
pp
```

Adding the data:

```{r}
pp + 
geom_point(aes(x = p, y = sort(pnorm(fheight, m, s))), data = (father.son))

```

The PP plot is also less sensitive to deviations in the tails.

A compromise between the QQ and PP plots uses the arcsine square root variance-stabilizing transformation, which makes the variability approximately constant across the range of the plot:

```{r}
vpp <- ggplot() +
geom_line(aes(x = asin(sqrt(p)), y = asin(sqrt(pnorm(x, m, s))), group = sim),
          color = "gray", data = gb)
vpp
```

Adding the data:


```{r}
vpp +
geom_point(aes(x = asin(sqrt(p)), y = sort(asin(sqrt(pnorm(fheight, m, s))))),
           data = (father.son))
```

## Plots For Assessing Model Fit
* Both QQ and PP plots can be used to asses how well a theoretical family of models fits your data, or your residuals.

* To use a PP plot you have to estimate the parameters first.

* For a location-scale family, like the normal distribution family, you can use a QQ plot with a standard member of the family.

* Some other families can use other transformations that lead to straight lines for family members:

The Weibull family is widely used in reliability modeling; its CDF is
$$F(t) = 1 - \exp\left\{-\left(\frac{t}{b}\right)^a\right\}$$

* The logarithms of Weibull random variables form a location-scale family.

* Special paper used to be available for Weibull probability plots.

A Weibull QQ plot for price in the diamonds data:

```{r}
n <- nrow(diamonds)
p <- (1 : n) / n - 0.5 / n
ggplot(diamonds) +
    geom_point(aes(x = log10(qweibull(p, 1, 1)), y = log10(sort(price))))
```


* The lower tail does not match a Weibull distribution.

* Is this important?

* In engineering applications it often is.

* In selecting a reasonable model to capture the shape of this distribution it may not be.

* QQ plots are helpful for understanding departures from a theoretical model.

* No data will fit a theoretical model perfectly.

* Case-specific judgment is needed to decide whether departures are important.

* George Box: All models are wrong but some are useful.
