# Visualizing residuals

Source: https://www.r-bloggers.com/visualising-residuals/

```{r}
fit <- lm(mpg ~ hp, data = mtcars)  # Fit the model
summary(fit)  # Report the results
```



```{r}
par(mfrow = c(2, 2))  # Split the plotting panel into a 2 x 2 grid
plot(fit)  # Plot the model information

par(mfrow = c(1, 1))  # Return plotting panel to 1 section
```

## Simple Linear Regression

```{r}
d <- mtcars
fit <- lm(mpg ~ hp, data = d)
```

```{r}
d$predicted <- predict(fit)   # Save the predicted values
d$residuals <- residuals(fit) # Save the residual values

# Quick look at the actual, predicted, and residual values
library(dplyr)
d %>% select(mpg, predicted, residuals) %>% head()

```


### Step 3: plot the actual and predicted values

> plot first the actual data

```{r}
library(ggplot2)
ggplot(d, aes(x = hp, y = mpg)) +  # Set up canvas with outcome variable on y-axis
  geom_point()  # Plot the actual points
```

> Next, we plot the predicted values in a way that they’re distinguishable from the actual values. For example, let’s change their shape:

```{r}
ggplot(d, aes(x = hp, y = mpg)) +
  geom_point() +
  geom_point(aes(y = predicted), shape = 1)  # Add the predicted values
```

> This is on track, but it’s difficult to see how our actual and predicted values are related. Let’s connect the actual data points with their corresponding predicted value using geom_segment():

```{r}
ggplot(d, aes(x = hp, y = mpg)) +
  geom_segment(aes(xend = hp, yend = predicted)) +
  geom_point() +
  geom_point(aes(y = predicted), shape = 1)
```

> We’ll make a few final adjustments:
* Clean up the overall look with theme_bw().
* Fade out connection lines by adjusting their alpha.
* Add the regression slope with geom_smooth():

```{r}

library(ggplot2)
ggplot(d, aes(x = hp, y = mpg)) +
  geom_smooth(method = "lm", se = FALSE, color = "lightgrey") +  # Plot regression slope
  geom_segment(aes(xend = hp, yend = predicted), alpha = .2) +  # alpha to fade lines
  geom_point() +
  geom_point(aes(y = predicted), shape = 1) +
  theme_bw()  # Add theme for cleaner look
```


## Step 4: use residuals to adjust

> Finally, we want to make an adjustment to highlight the size of the residual. There are MANY options. To make comparisons easy, I’ll make adjustments to the actual values, but you could just as easily apply these, or other changes, to the predicted values. Here are a few examples building on the previous plot:

```{r}


# ALPHA
# Changing alpha of actual values based on absolute value of residuals
ggplot(d, aes(x = hp, y = mpg)) +
  geom_smooth(method = "lm", se = FALSE, color = "lightgrey") +
  geom_segment(aes(xend = hp, yend = predicted), alpha = .2) +

  # > Alpha adjustments made here...
  geom_point(aes(alpha = abs(residuals))) +  # Alpha mapped to abs(residuals)
  guides(alpha = FALSE) +  # Alpha legend removed
  # <

  geom_point(aes(y = predicted), shape = 1) +
  theme_bw()
```


```{r}
# COLOR
# High residuals (in abolsute terms) made more red on actual values.
ggplot(d, aes(x = hp, y = mpg)) +
  geom_smooth(method = "lm", se = FALSE, color = "lightgrey") +
  geom_segment(aes(xend = hp, yend = predicted), alpha = .2) +

  # > Color adjustments made here...
  geom_point(aes(color = abs(residuals))) + # Color mapped to abs(residuals)
  scale_color_continuous(low = "black", high = "red") +  # Colors to use here
  guides(color = FALSE) +  # Color legend removed
  # <

  geom_point(aes(y = predicted), shape = 1) +
  theme_bw()

```


```{r}
# SIZE AND COLOR
# Same coloring as above, size corresponding as well
ggplot(d, aes(x = hp, y = mpg)) +
  geom_smooth(method = "lm", se = FALSE, color = "lightgrey") +
  geom_segment(aes(xend = hp, yend = predicted), alpha = .2) +

  # > Color AND size adjustments made here...
  geom_point(aes(color = abs(residuals), size = abs(residuals))) + # size also mapped
  scale_color_continuous(low = "black", high = "red") +
  guides(color = FALSE, size = FALSE) +  # Size legend also removed
  # <

  geom_point(aes(y = predicted), shape = 1) +
  theme_bw()
```



```{r}
# COLOR UNDER/OVER
# Color mapped to residual with sign taken into account.
# i.e., whether actual value is greater or less than predicted
ggplot(d, aes(x = hp, y = mpg)) +
  geom_smooth(method = "lm", se = FALSE, color = "lightgrey") +
  geom_segment(aes(xend = hp, yend = predicted), alpha = .2) +

  # > Color adjustments made here...
  geom_point(aes(color = residuals)) +  # Color mapped here
  scale_color_gradient2(low = "blue", mid = "white", high = "red") +  # Colors to use here
  guides(color = FALSE) +
  # <

  geom_point(aes(y = predicted), shape = 1) +
  theme_bw()
```

I particularly like this last example, because the colours nicely help to identify non-linearity in the data. For example, we can see that there is more red for extreme values of hp where the actual values are greater than what is being predicted. There is more blue in the centre, however, indicating that the actual values are less than what is being predicted. Together, this suggests that the relationship between the variables is non-linear, and might be better modelled by including a quadratic term in the regression equation.
